\thetitle

### 2015-11-26 compared.tex
telescopic CWFs

\begin{defn}
A functor $F:\cat{C}\to\cat{D}$ is said to
\emph{lift factorizations} if for any $f\in\cat{C}/X$, the functor
$F_{\mathbf{fact}(f)}:\mathbf{fact}(f)\to\mathbf{fact}(F(f))$ is an isomorphism
of categories. 
\end{defn}

@@

\begin{defn}
A category $\cat{C}$ with terminal object is said to be \emph{stratified} if there exists a 
\emph{stratification functor}
\begin{equation*}
L : \cat{C}\to (\mathbb{N},\geq)
\end{equation*}
such that
\begin{enumerate}
\item $L(X)\jdeq 0$ if and only if $X$ is terminal, and for any $f:X\to Y$ we have
$L(X)\jdeq L(Y)$ if and only if $X\jdeq Y$ and $f\jdeq\catid{X}$. 
\item every morphism $f:X\to Y$ in $\cat{C}_\mathbb{E}$, where $L(X)\jdeq
n+m+1$ and $L(Y)\jdeq n$, has a unique factorization 
\begin{equation*}
\begin{tikzcd}
X \arrow[r,"f_m"] & X_m \arrow[r,"f_{m-1}"] & \cdots \arrow[r,"f_1"] & X_1 \arrow[r,"f_0"] & Y
\end{tikzcd}
\end{equation*}
where $L(X_i)\jdeq n+i$.
\end{enumerate}
A functor $F:\cat{C}\to\cat{D}$ between stratified categories is said to be stratified 
if $L_{\cat{C}}\jdeq L_{\cat{D}}\circ F$. 
\end{defn}

@@

\begin{defn}
A pre-E-system is said to be \emph{stratified} if its underlying category is
stratified and if each $W_A$ and $S_x$ is a stratified functor. The category of
stratified pre-E-systems with stratified E-homomorphisms between them is denoted by
$\mathbf{Esys_s}$.
\end{defn}

@@

\begin{defn}
A \emph{B-framework} is a collection of data of the following form:
\begin{enumerate}
\item for all $n\in\mathbb{N}$ two sets $B_n$ and $\tilde{B}_n$. 
\item for all $n\in\mathbb{N}$ maps of the form
\begin{align*}
\eft[n] & : B_{n+1}\to B_n \\
\ebd[n] & : \tilde{B}_n\to B_n.
\end{align*}
For $m,n\in\mathbb{N}$, we denote the composition $\eft[n]\circ\cdots\circ\eft[n+m]:B_{n+m+1}\to B_n$ by $\eft[n]^m$. 
\item $B_0$ is a singleton $\{\pt\}$.
\end{enumerate} 
\end{defn}

@@

\begin{defn}
A homomorphism $H:\mathbb{B}\to\mathbb{A}$ of B-frameworks consists of maps
$H_n:B_n\to A_n$ and $\tilde{H}_n:\tilde{B}_n\to\tilde{A}_n$ such that
\begin{align*}
\eft(H(X)) & \jdeq H(\eft(X)) \\
\ebd(\tilde{H}(x)) & \jdeq H(\ebd(x))
\end{align*}
for any $X\in B_n$ and $x\in\tilde{B}_n$. The category of B-frameworks is
denoted by $\mathbf{Bfr}$. 
\end{defn}

@@

\begin{defn}
A B0-system $\mathbb{B}$ consists of a B-framework $\mathbb{B}$ and homomorphisms
\begin{align*}
W_{X} & : \mathbb{B}/\eft(X)\to\mathbb{B}/X\\
S_{x} & : \mathbb{B}/\ebd(x)\to\mathbb{B}/\eft(\ebd(x))
\end{align*}
of B-frameworks, for any $X\in B_{n+1}$ and any $x\in\tilde{B}_{n+1}$. 
\end{defn}

@@

\begin{defn}
A B0-homomorphism $H:\mathbb{B}\to\mathbb{A}$ is a homomorphism of B-frameworks
for which the diagrams
\begin{equation*}
\begin{tikzcd}[column sep=huge]
\mathbb{B}/X \arrow[r,"H/X"] & \mathbb{A}/H(X) \\
\mathbb{B}/\eft(X) \arrow[u,"W_X"] \arrow[r,swap,"H/\eft(X)"] & \mathbb{A}/\eft(H(X)) \arrow[u,swap,"W_{H(X)}"]
\end{tikzcd}
\end{equation*}
and
\begin{equation*}
\begin{tikzcd}[column sep=huge]
\mathbb{B}/\ebd(x) \arrow[r,"H/\ebd(x)"] \arrow[d,swap,"S_x"] & \mathbb{A}/\ebd(H(X)) \arrow[d,"S_{\tilde{H}(x)}"] \\
\mathbb{B}/\eft(\ebd(x)) \arrow[r,swap,"H/\eft(\ebd(x))"] & \mathbb{A}/\eft(\ebd(H(X)))
\end{tikzcd}
\end{equation*}
of homomorphisms of B-frameworks commute for every 
$X\in B_n$ and $x\in\tilde{B}_n$. The category of B0-systems is denoted
by $\mathbf{B0sys}$.
\end{defn}

@@

\begin{defn}
A unital B0-system is a B0-system with maps
\begin{equation*}
\delta_n : B_{n+1}\to \tilde{B}_{n+2}.
\end{equation*}
such that $\ebd(\delta_n(X))\jdeq W_{X}(X)$ for any $X\in B_n$. 
A B0-homomorphism $H$ is said to be unital if one has
\begin{equation*}
H(\delta(X))\jdeq\delta(H(X))
\end{equation*}
for any $X\in B_n$.  
\end{defn}

@@

\subsection{Comprehension categories and categories with families}
In this section we recall Dybjer's notion of categories with families and we
introduce the equivalent notion of D-systems. After having proved that categories with families
are the same things as D-systems, we will focus on constructing a canonical functor from
D-systems to E'-systems, in which the category $\cat{F}$ of families has contexts
as objects and compositions of projections as morphisms. There is no reason why
this category should possess a terminal object, so we naturally have to retreat
to E'-systems without empty contex. However, the construction of a
canonical functor in which families are represented by compositions of projections
is still not possible without a further coherence condition. To make sure that
the pullback operations of a D-system extend to the category
$\cat{F}$, we introduce \emph{coherent} D-systems. So we only have a functor $U$
from the category $\mathbf{Dsys}$ of coherent D-systems to the category
$\mathbf{E'sys_{\circ}}$ of E'-systems without empty context.
We see that it is not evident at all to turn categories with families into
E-systems, even if we ignore the issue that categories with families map
naturally into E'-systems without empty context.
Thus it seems that the type theories of categories with families and of E-systems
are not directly comparable.

\begin{defn}
Let $p:\cat{E}\to\cat{B}$ be a functor and let $f:\Delta\to \Gamma$ be a morphism in $\cat{B}$.
A morphism $g:A\to B$ in $\cat{E}$ is said to be \emph{cartesian over $f$}, if
\begin{enumerate}
\item $p(g)\jdeq f$, and 
\item For all $g':A'\to B$ with $p(g')\jdeq f$, there is a unique $u:A'\to A$
such that $g'=g\circ u$ and $p(u)\jdeq\catid{\Delta}$. 
\end{enumerate}
Dually, a morphism $g:A\to B$ in $\cat{E}$ is said to be \emph{cocartesian over
$f$}, if
\begin{enumerate}
\item $p(g)\jdeq f$, and
\item For all $g':A\to B'$ with $p(g')\jdeq f$, there is a unique $v:B\to B'$
such that $g'=v\circ g$ and $p(v)\jdeq\catid{\Gamma}$. 
\end{enumerate}
A (co)cartesian morphism over $f$ is also called a \emph{(co)cartesian lift of $f$}.
\end{defn}

\begin{defn}
A functor $p:\cat{E}\to\cat{B}$ is called a \emph{Grothendieck fibration} if
every morphism into an object of the form $p(A)$ for some object $A$ of $\cat{E}$ 
has a cartesian lift, and cartesian morphisms are closed under composition.
\end{defn}

\begin{defn}
A Grothendieck fibration $p:\cat{E}\to\cat{B}$ is said to be \emph{split}, if
there is a functorial choice of cartesian lifts, i.e.~for every $f:\Delta\to \Gamma$ in 
$\cat{B}$ and any $B\in \cat{E}$ such that $p(B)\jdeq \Delta$,
there is a cartesian lift $j(B,g):A \to B$ for some object $A$ of $\cat{E}$, such that 
$j(B,p(B))\jdeq\catid{B}$ and $j(B, g\circ h)\jdeq j(B,g)\circ j(\mathrm{dom}(j(B,g)),h)$.
\end{defn}

\begin{defn}
A comprehension category $\mathcal{P}$ consists of categories $\cat{E}$, $\cat{C}$ and a
commuting triangle
\begin{equation*}
\begin{tikzcd}
\cat{E} \arrow[rr,"\mathcal{P}"] \arrow[dr,swap,"p"] & & \cat{C}^{\to} \arrow[dl,"\mathrm{cod}"] \\
& \cat{C}
\end{tikzcd}
\end{equation*}
of functors, such that $p$ is a Grothendieck fibration, and $\mathcal{P}$ maps
cartesian morphisms in $\cat{E}$ to pullback squares in $\cat{C}$. We say that
$\mathcal{P}$ is a full comprehension category if the functor $\mathcal{P}:
\cat{E}\to\cat{C}^{\to}$ is full and faithful; and we say that
$\mathcal{P}$ is split if $p$ is split.
\end{defn}

We now recall the notion of categories with families, which forms a class of
examples of comprehension categories.

\begin{defn}
A \emph{category with families} consists of
\begin{enumerate}
\item A category $\cat{C}$ with a terminal object $1$.
\item A presheaf $\mathrm{Ty}:\op{\cat{C}}\to\mathbf{Set}$. We will write
$f_\ast(A)$ for $\mathrm{Ty}(f,A)$.
\item For any $A\in\mathrm{Ty}(\Gamma)$, a morphism
\begin{equation*}
p_A : \ctxext{\Gamma}{A}\twoheadrightarrow \Gamma
\end{equation*}
in $\cat{C}$.
\item a preasheaf $\mathrm{Tm}:\op{(\int_\cat{C}\mathrm{Ty})}\to\mathbf{Set}$.
The action of a morphism $f:(\Delta,f_\ast(A))\to(\Gamma,A)$ of $\int_\cat{C}\mathrm{Ty}$
on a term $x\in \mathrm{Tm}(A)$ is also written by $f_\ast(x)$.
\item a term $\delta_A\in\mathrm{Tm}(\ctxext{\Gamma}{A},(p_A)_\ast(A))$ for every
$A\in\mathrm{Ty}(\Gamma)$, 
\end{enumerate}
with the universal property that for every $f:\Delta\to\Gamma$ in $\cat{C}$, 
the function
\begin{equation*}
\theta\mapsto\theta_\ast(\delta_A)
  : \{\theta:\Delta\to\ctxext{\Gamma}{A}\mid p_A\circ\theta\jdeq f\}\to
    \mathrm{Tm}(f_\ast(A))
\end{equation*}
is a bijection. We denote the inverse of this map by $\mu_{A,f}$. 
\end{defn}

@@

There is essentially only one choice for the presheaf $\mathrm{Tm}$ and the
term $\delta_A$, so the extra structure on a setup for categories with families
is determined essentially uniquely once it exists. In other words, being a
category with families is a property on setups for categories with families.

@@

\begin{defn}
A \emph{D-system} (a category with attributes, see \cite{Hofmann_syntax_semantics}) consists of
\begin{enumerate}
\item A category $\cat{C}$ with a terminal object $1$.
\item A presheaf $\mathrm{Ty}:\op{\cat{C}}\to\mathbf{Set}$. 
\item A commuting triangle
\begin{equation*}
\begin{tikzcd}
\int_{\cat{C}}\mathrm{Ty} \arrow[rr,"p"] \arrow[dr,"\pi"] & & \cat{C}^\square \arrow[dl,"\mathrm{cod}"] \\
& \cat{C}
\end{tikzcd}
\end{equation*}
of functors, where $\cat{C}^\square$ is the sub-pre-category of $\cat{C}^{\to}$ of
which the morphisms are pullback squares. For any $A\in\mathrm{Ty}(\Gamma)$,
the morphism $p(\Gamma,A)$ will be denoted by $p_A$; for any
$(f,A):(\Delta,f_\ast(A))\to(\Gamma,A)$ in $\int_{\cat{C}}\mathrm{Ty}$, the top
morphism of the pullback square $p(f,A)$ will be denoted by $\pi_2(f,A)$.
\end{enumerate}
\end{defn}

@@

\begin{defn}
A \emph{D-homomorphism from $\mathbb{D}\to\mathbb{D}'$} consists of a functor 
$\mathcal{F}:\cat{C}\to\cat{C}'$ which maps the terminal object of
$\cat{C}$ to the terminal object of $\cat{C}'$, and a natural transformation
$\tau:\mathrm{Ty}\Rightarrow\mathrm{Ty}'\circ\mathcal{F}$, such that 
$\mathcal{F}(p_A)\jdeq p_{\mathcal{F}(A)}$ for any
$A\in\mathrm{Ty}(\Gamma)$, and $\mathcal{F}(\pi_2(f,A))\jdeq
\pi_2(\mathcal{F}(f),\tau_\Gamma(A))$ for any $f:\Delta\to\Gamma$ in $\cat{C}$ and
any $A\in\mathrm{Ty}(\Gamma)$. 
\end{defn}

@@

\begin{thm}\label{thm:DD}
The category of categories with families is equivalent to the category of 
D-systems.\hfill$\square$
\end{thm}

\begin{comment}
Since this theorem basically asserts that a category with families is a
comprehension category

\begin{proof}[Outline of the proof]
Let $\mathbf{CWF}$ be a category with families. The fact that $\mathbf{CWF}$ is
a D-system can be shown by proving each of the assertions in the following six 
steps.
\begin{enumerate}
\item For every $f:\Delta\to\Gamma$ and $A\in\mathrm{Ty}(\Gamma)$, there is an
isomorphism
\begin{equation*}
\zeta_{A,f}:\{\theta:\Delta\to\ctxext{\Gamma}{A}\mid p_A\circ\theta\jdeq f\}
  \cong
\{\eta:\Delta\to\ctxext{\Delta}{f_\ast(A)}\mid p_{f_\ast(A)}\circ\eta\jdeq\catid{\Delta}\}
\end{equation*}
natural in $f$ in the sense that the equalities
\begin{align*}
\zeta_{A,\catid{\Gamma}}(\theta) & \jdeq \theta \\
\zeta_{A,f\circ g}(\theta\circ g) & \jdeq \zeta_{f_\ast(A),g}(\zeta_{A,f}(\theta)\circ g)
\end{align*}
hold for any $f:\Delta\to\Gamma$ and $g:X\to\Delta$, and any $\theta:\Delta\to\ctxext{\Gamma}{A}$
such that $p_A\circ\theta\jdeq f$.
Furthermore, we get $\zeta_{A,p_A}(\catid{\ctxext{\Gamma}{A}})\jdeq\varphi_{(p_A)_\ast(A)}(\delta_A)$.

\medskip
(Proof omitted).
\begin{comment}
By \autoref{lem:cwf_to_dsys} we may define
\begin{equation*}
\zeta_{A,f}(\theta)\defeq \varphi_{f_\ast(A)}(\theta_\ast(\delta_A))
\end{equation*}
for any $\theta\in\{\theta:\Delta\to\ctxext{\Gamma}{A}\mid p_A\circ\theta\jdeq f\}$. 
Since $\zeta_{A,f}$ is a composition of isomorphism, it is itself an isomorphism.
Also, we get immediately that $\zeta_{A,\catid{\Gamma}}(\theta)\jdeq\theta$. To
show the remaining identity:
\begin{align*}
\zeta_{A,f\circ g}(\theta\circ g)
& \jdeq \varphi_{g_\ast(f_\ast(A))}((\theta\circ g)_\ast(\delta_A)) \\
& \jdeq \varphi_{g_\ast(f_\ast(A))}(g_\ast(\theta_\ast(\delta_A))) \\
& \jdeq \varphi_{g_\ast(f_\ast(A))}(g_\ast(\varphi_{f_\ast(A)}(\theta_\ast(\delta_A))_\ast(\delta_A))) \\
& \jdeq \varphi_{g_\ast(f_\ast(A))}(g_\ast(\zeta_{A,f}(\theta)_\ast(\delta_A))) \\
& \jdeq \varphi_{g_\ast(f_\ast(A))}((\zeta_{A,f}(\theta)\circ g)_\ast(\delta_A)) \\
& \jdeq \zeta_{f_\ast(A),g}(\zeta_{A,f}(\theta)\circ g).
\end{align*}
This completes the proof that the isomorphisms $\zeta_{A,f}$ satisfy the required
equalities.
\end{comment}

\begin{comment}
\item
For any $A\in\mathrm{Ty}(\Gamma)$, we define
$W_A\defeq (p_A)_\ast:\cat{F}/\Gamma\to\cat{F}/\ctxext{\Gamma}{A}$. Then we get
for any $A,B\in\cat{F}/\Gamma$ an isomorphism
\begin{equation*}
\mathrm{hom}_{\cat{C}/\Gamma}(A,B)
  \cong
T(W_A(B)).
\end{equation*}
In particular, we get a term $\idtm{A}\in T((p_A)_\ast(A))$. 
\medskip

@@

\medskip
(Proof omitted).
\begin{comment}
Note that we have the isomorphism
\begin{align*}
& \{ \theta : \ctxext{\Delta}{f_\ast(A)}\to\ctxext{\Gamma}{A}\mid p_A\circ\theta\jdeq f\circ p_{f_\ast(A)} \} \\
  & \cong
\{ \eta : \ctxext{\Delta}{f_\ast(A)}\to\ctxext{{\Delta}{f_\ast(A)}}{(f\circ p_{f_\ast(A)})_\ast(A)}
  \mid (f\circ p_{f_\ast(A)})_\ast(A)\circ\eta\jdeq\catid{\ctxext{\Delta}{f_\ast(A)}}\} \\
  & \jdeq
\{ \eta : \ctxext{\Delta}{f_\ast(A)}\to\ctxext{{\Delta}{f_\ast(A)}}{(p_{f_\ast(A)})_\ast(f_\ast(A))}
  \mid (f\circ p_{f_\ast(A)})_\ast(A)\circ\eta\jdeq\catid{\ctxext{\Delta}{f_\ast(A)}}\} \\
  & \jdeq
T((p_{f_\ast(A)})_\ast(f_\ast(A))).
\end{align*}
Since we have $\delta_{f_\ast(A)}\in T((p_{f_\ast(A)})_\ast(f_\ast(A)))$, we obtain 
\begin{equation*}
\pi_2(f,A)\defeq \zeta_{A,f\circ p_{f_\ast(A)}}^{-1}(\delta_{f_\ast(A)}):\ctxext{\Delta}{f_\ast(A)}\to\ctxext{\Gamma}{A}
\end{equation*}
satisfying $p_A\circ\pi_2(f,A)\jdeq f\circ p_{f_\ast(A)}$.
\end{comment}
\begin{comment}
\item For any $f:\Delta\to\Gamma$, $g:X\to\Delta$ and $A\in\mathrm{Ty}(\Gamma)$, one has
\begin{equation*}
\pi_2(f,A)\circ\pi_2(g,f_\ast(A))\jdeq\pi_2(f\circ g,A).
\end{equation*}

>>>>

\begin{defn}
A functor $F:\cat{C}\to\cat{D}$ is said to
\define{lift factorizations} if for any $f\in\cat{C}/X$, the functor
$F_{\mathbf{fact}(f)}:\mathbf{fact}(f)\to\mathbf{fact}(F(f))$ is an isomorphism
of categories. 
\end{defn}

@@

\begin{defn}
A category $\cat{C}$ with terminal object is said to be \define{stratified} if there exists a 
\emph{stratification functor}
\begin{equation*}
L : \cat{C}\to (\mathbb{N},\geq)
\end{equation*}
such that
\begin{enumerate}
\item $L(X)\jdeq 0$ if and only if $X$ is terminal, and for any $f:X\to Y$ we have
$L(X)\jdeq L(Y)$ if and only if $X\jdeq Y$ and $f\jdeq\catid{X}$. 
\item every morphism $f:X\to Y$ in $\cat{C}_\mathbb{E}$, where $L(X)\jdeq
n+m+1$ and $L(Y)\jdeq n$, has a unique factorization 
\begin{equation*}
\begin{tikzcd}
X \arrow[r,"f_m"] & X_m \arrow[r,"f_{m-1}"] & \cdots \arrow[r,"f_1"] & X_1 \arrow[r,"f_0"] & Y
\end{tikzcd}
\end{equation*}
where $L(X_i)\jdeq n+i$.
\end{enumerate}
A functor $F:\cat{C}\to\cat{D}$ between stratified categories is said to be stratified 
if $L_{\cat{C}}\jdeq L_{\cat{D}}\circ F$. 
\end{defn}

@@

\begin{defn}
A pre-E-system is said to be \define{stratified} if its underlying category is
stratified and if each $W_A$ and $S_x$ is a stratified functor. The category of
stratified pre-E-systems with stratified E-homomorphisms between them is denoted by
$\mathbf{Esys_s}$.
\end{defn}

@@

\begin{defn}
A \define{B-framework} is a collection of data of the following form:
\begin{enumerate}
\item for all $n\in\mathbb{N}$ two sets $B_n$ and $\tilde{B}_n$. 
\item for all $n\in\mathbb{N}$ maps of the form
\begin{align*}
\eft[n] & : B_{n+1}\to B_n \\
\ebd[n] & : \tilde{B}_n\to B_n.
\end{align*}
For $m,n\in\mathbb{N}$, we denote the composition $\eft[n]\circ\cdots\circ\eft[n+m]:B_{n+m+1}\to B_n$ by $\eft[n]^m$. 
\item $B_0$ is a singleton $\{\pt\}$.
\end{enumerate} 
\end{defn}

@@

\begin{defn}
A \define{homomorphism $H:\mathbb{B}\to\mathbb{A}$ of B-frameworks} consists of maps
$H_n:B_n\to A_n$ and $\tilde{H}_n:\tilde{B}_n\to\tilde{A}_n$ such that
\begin{align*}
\eft(H(X)) & \jdeq H(\eft(X)) \\
\ebd(\tilde{H}(x)) & \jdeq H(\ebd(x))
\end{align*}
for any $X\in B_n$ and $x\in\tilde{B}_n$. The category of B-frameworks is
denoted by $\mathbf{Bfr}$. 
\end{defn}

@@

\begin{defn}
A \define{B0-system} $\mathbb{B}$ consists of a B-framework $\mathbb{B}$ and homomorphisms
\begin{align*}
W_{X} & : \mathbb{B}/\eft(X)\to\mathbb{B}/X\\
S_{x} & : \mathbb{B}/\ebd(x)\to\mathbb{B}/\eft(\ebd(x))
\end{align*}
of B-frameworks, for any $X\in B_{n+1}$ and any $x\in\tilde{B}_{n+1}$. 
\end{defn}

@@

\begin{defn}
A \define{B0-homomorphism} $H:\mathbb{B}\to\mathbb{A}$ is a homomorphism of B-frameworks
for which the diagrams
\begin{equation*}
\begin{tikzcd}[column sep=huge]
\mathbb{B}/X \arrow[r,"H/X"] & \mathbb{A}/H(X) \\
\mathbb{B}/\eft(X) \arrow[u,"W_X"] \arrow[r,swap,"H/\eft(X)"] & \mathbb{A}/\eft(H(X)) \arrow[u,swap,"W_{H(X)}"]
\end{tikzcd}
\end{equation*}
and
\begin{equation*}
\begin{tikzcd}[column sep=huge]
\mathbb{B}/\ebd(x) \arrow[r,"H/\ebd(x)"] \arrow[d,swap,"S_x"] & \mathbb{A}/\ebd(H(X)) \arrow[d,"S_{\tilde{H}(x)}"] \\
\mathbb{B}/\eft(\ebd(x)) \arrow[r,swap,"H/\eft(\ebd(x))"] & \mathbb{A}/\eft(\ebd(H(X)))
\end{tikzcd}
\end{equation*}
of homomorphisms of B-frameworks commute for every 
$X\in B_n$ and $x\in\tilde{B}_n$. The category of B0-systems is denoted
by $\mathbf{B0sys}$.
\end{defn}

@@

\begin{defn}
A \define{unital B0-system} is a B0-system with maps
\begin{equation*}
\delta_n : B_{n+1}\to \tilde{B}_{n+2}.
\end{equation*}
such that $\ebd(\delta_n(X))\jdeq W_{X}(X)$ for any $X\in B_n$. 
A B0-homomorphism $H$ is said to be \define{unital} if one has
\begin{equation*}
H(\delta(X))\jdeq\delta(H(X))
\end{equation*}
for any $X\in B_n$.  
\end{defn}

@@

\subsection{Telescopic categories with families}

\begin{defn}
A \define{category with families} consists of
\begin{enumerate}
\item A category $\cat{C}$ with a terminal object $1$.
\item A presheaf $\mathrm{Ty}:\op{\cat{C}}\to\mathbf{Set}$. We will write
$f_\ast(A)$ for $\mathrm{Ty}(f,A)$.
\item For any $A\in\mathrm{Ty}(\Gamma)$, a morphism
\begin{equation*}
p_A : \ctxext{\Gamma}{A}\twoheadrightarrow \Gamma
\end{equation*}
in $\cat{C}$.
\item a preasheaf $\mathrm{Tm}:\op{(\int_\cat{C}\mathrm{Ty})}\to\mathbf{Set}$.
The action of a morphism $f:(\Delta,f_\ast(A))\to(\Gamma,A)$ of $\int_\cat{C}\mathrm{Ty}$
on a term $x\in \mathrm{Tm}(A)$ is also written by $f_\ast(x)$.
\item a term $\delta_A\in\mathrm{Tm}(\ctxext{\Gamma}{A},(p_A)_\ast(A))$ for every
$A\in\mathrm{Ty}(\Gamma)$, 
\end{enumerate}
with the universal property that for every $f:\Delta\to\Gamma$ in $\cat{C}$, 
the function
\begin{equation*}
\theta\mapsto\theta_\ast(\delta_A)
  : \{\theta:\Delta\to\ctxext{\Gamma}{A}\mid p_A\circ\theta\jdeq f\}\to
    \mathrm{Tm}(f_\ast(A))
\end{equation*}
is a bijection. We denote the inverse of this map by $\mu_{A,f}$. 
\end{defn}

@@

There is essentially only one choice for the presheaf $\mathrm{Tm}$ and the
term $\delta_A$, so the extra structure on a setup for categories with families
is determined essentially uniquely once it exists.

@@

\begin{defn}
A \define{D-system} (a category with attributes, see \cite{Hofmann_syntax_semantics}) consists of
\begin{enumerate}
\item A category $\cat{C}$ with a terminal object $1$.
\item A presheaf $\mathrm{Ty}:\op{\cat{C}}\to\mathbf{Set}$. 
\item A commuting triangle
\begin{equation*}
\begin{tikzcd}
\int_{\cat{C}}\mathrm{Ty} \arrow[rr,"p"] \arrow[dr,"\pi"] & & \cat{C}^\square \arrow[dl,"\mathrm{cod}"] \\
& \cat{C}
\end{tikzcd}
\end{equation*}
of functors, where $\cat{C}^\square$ is the sub-pre-category of $\cat{C}^{\to}$ of
which the morphisms are pullback squares. For any $A\in\mathrm{Ty}(\Gamma)$,
the morphism $p(\Gamma,A)$ will be denoted by $p_A$; for any
$(f,A):(\Delta,f_\ast(A))\to(\Gamma,A)$ in $\int_{\cat{C}}\mathrm{Ty}$, the top
morphism of the pullback square $p(f,A)$ will be denoted by $\pi_2(f,A)$.
\end{enumerate}
\end{defn}

@@

\begin{defn}
A \define{D-homomorphism} from $\mathbb{D}$ to $\mathbb{D}'$ consists of a functor 
$\mathcal{F}:\cat{C}\to\cat{C}'$ which maps the terminal object of
$\cat{C}$ to the terminal object of $\cat{C}'$, and a natural transformation
$\tau:\mathrm{Ty}\Rightarrow\mathrm{Ty}'\circ\mathcal{F}$, such that 
$\mathcal{F}(p_A)\jdeq p_{\mathcal{F}(A)}$ for any
$A\in\mathrm{Ty}(\Gamma)$, and $\mathcal{F}(\pi_2(f,A))\jdeq
\pi_2(\mathcal{F}(f),\tau_\Gamma(A))$ for any $f:\Delta\to\Gamma$ in $\cat{C}$ and
any $A\in\mathrm{Ty}(\Gamma)$. 
\end{defn}

@@

\begin{thm}\label{thm:DD}
The category of categories with families is equivalent to the category of 
D-systems.%\hfill$\square$
\end{thm}

\begin{comment}
Since this theorem basically asserts that a category with families is a
comprehension category
\end{comment}

\begin{proof}%[Outline of the proof]
Let $\mathbf{CWF}$ be a category with families. The fact that $\mathbf{CWF}$ is
a D-system can be shown by proving each of the assertions in the following six 
steps.
\begin{enumerate}
\item For every $f:\Delta\to\Gamma$ and $A\in\mathrm{Ty}(\Gamma)$, there is an
isomorphism
\begin{equation*}
\zeta_{A,f}:\{\theta:\Delta\to\ctxext{\Gamma}{A}\mid p_A\circ\theta\jdeq f\}
  \cong
\{\eta:\Delta\to\ctxext{\Delta}{f_\ast(A)}\mid p_{f_\ast(A)}\circ\eta\jdeq\catid{\Delta}\}
\end{equation*}
natural in $f$ in the sense that the equalities
\begin{align*}
\zeta_{A,\catid{\Gamma}}(\theta) & \jdeq \theta \\
\zeta_{A,f\circ g}(\theta\circ g) & \jdeq \zeta_{f_\ast(A),g}(\zeta_{A,f}(\theta)\circ g)
\end{align*}
hold for any $f:\Delta\to\Gamma$ and $g:X\to\Delta$, and any $\theta:\Delta\to\ctxext{\Gamma}{A}$
such that $p_A\circ\theta\jdeq f$.
Furthermore, we get $\zeta_{A,p_A}(\catid{\ctxext{\Gamma}{A}})\jdeq\varphi_{(p_A)_\ast(A)}(\delta_A)$.

\medskip
%(Proof omitted).
%\begin{comment}
By \autoref{lem:cwf_to_dsys} we may define
\begin{equation*}
\zeta_{A,f}(\theta)\defeq \varphi_{f_\ast(A)}(\theta_\ast(\delta_A))
\end{equation*}
for any $\theta\in\{\theta:\Delta\to\ctxext{\Gamma}{A}\mid p_A\circ\theta\jdeq f\}$. 
Since $\zeta_{A,f}$ is a composition of isomorphism, it is itself an isomorphism.
Also, we get immediately that $\zeta_{A,\catid{\Gamma}}(\theta)\jdeq\theta$. To
show the remaining identity:
\begin{align*}
\zeta_{A,f\circ g}(\theta\circ g)
& \jdeq \varphi_{g_\ast(f_\ast(A))}((\theta\circ g)_\ast(\delta_A)) \\
& \jdeq \varphi_{g_\ast(f_\ast(A))}(g_\ast(\theta_\ast(\delta_A))) \\
& \jdeq \varphi_{g_\ast(f_\ast(A))}(g_\ast(\varphi_{f_\ast(A)}(\theta_\ast(\delta_A))_\ast(\delta_A))) \\
& \jdeq \varphi_{g_\ast(f_\ast(A))}(g_\ast(\zeta_{A,f}(\theta)_\ast(\delta_A))) \\
& \jdeq \varphi_{g_\ast(f_\ast(A))}((\zeta_{A,f}(\theta)\circ g)_\ast(\delta_A)) \\
& \jdeq \zeta_{f_\ast(A),g}(\zeta_{A,f}(\theta)\circ g).
\end{align*}
This completes the proof that the isomorphisms $\zeta_{A,f}$ satisfy the required
equalities.
%\end{comment}

%\begin{comment}
\item
For any $A\in\mathrm{Ty}(\Gamma)$, we define
$W_A\defeq (p_A)_\ast:\cat{F}/\Gamma\to\cat{F}/\ctxext{\Gamma}{A}$. Then we get
for any $A,B\in\cat{F}/\Gamma$ an isomorphism
\begin{equation*}
\mathrm{hom}_{\cat{C}/\Gamma}(A,B)
  \cong
T(W_A(B)).
\end{equation*}
In particular, we get a term $\idtm{A}\in T((p_A)_\ast(A))$. 
\medskip

@@

\medskip
%(Proof omitted).
%\begin{comment}
Note that we have the isomorphism
\begin{align*}
& \{ \theta : \ctxext{\Delta}{f_\ast(A)}\to\ctxext{\Gamma}{A}\mid p_A\circ\theta\jdeq f\circ p_{f_\ast(A)} \} \\
  & \cong
\{ \eta : \ctxext{\Delta}{f_\ast(A)}\to\ctxext{{\Delta}{f_\ast(A)}}{(f\circ p_{f_\ast(A)})_\ast(A)}
  \mid (f\circ p_{f_\ast(A)})_\ast(A)\circ\eta\jdeq\catid{\ctxext{\Delta}{f_\ast(A)}}\} \\
  & \jdeq
\{ \eta : \ctxext{\Delta}{f_\ast(A)}\to\ctxext{{\Delta}{f_\ast(A)}}{(p_{f_\ast(A)})_\ast(f_\ast(A))}
  \mid (f\circ p_{f_\ast(A)})_\ast(A)\circ\eta\jdeq\catid{\ctxext{\Delta}{f_\ast(A)}}\} \\
  & \jdeq
T((p_{f_\ast(A)})_\ast(f_\ast(A))).
\end{align*}
Since we have $\delta_{f_\ast(A)}\in T((p_{f_\ast(A)})_\ast(f_\ast(A)))$, we obtain 
\begin{equation*}
\pi_2(f,A)\defeq \zeta_{A,f\circ p_{f_\ast(A)}}^{-1}(\delta_{f_\ast(A)}):\ctxext{\Delta}{f_\ast(A)}\to\ctxext{\Gamma}{A}
\end{equation*}
satisfying $p_A\circ\pi_2(f,A)\jdeq f\circ p_{f_\ast(A)}$.
%\end{comment}
%\begin{comment}
\item For any $f:\Delta\to\Gamma$, $g:X\to\Delta$ and $A\in\mathrm{Ty}(\Gamma)$, one has
\begin{equation*}
\pi_2(f,A)\circ\pi_2(g,f_\ast(A))\jdeq\pi_2(f\circ g,A).
\end{equation*}
